Avoiding the Global Sort: \break A Faster Contour Tree Algorithm

### 2017-04-15 paint-contour-tree.tex
various small  changes

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, with Morse degree at most $3$. We define this formally 
later, but it suffices to think of this as a general position assumption.
The degree assumption is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\subsection{Previous Work}

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all vertices of $\MM$ have distinct function values, 
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

@@

\subsection{Some remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees.
\Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

@@

The intuition is that the paint spilling from the maxmima in the simplicial complex, corresponds to paint spilling from the leaves in 
the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
%
This non-contiguous complication along with the fact that heap sizes keep changing as color classes merge, causes the analysis to be technically challenging. 
We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.

@@

We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
complex $\MM$.  In subsequent sections we then show how to reduce computing the contour to computing join by using contour surgery.

@@

\begin{compactenum}
%     \item Construct a painting of $\MM$ by a descending BFS 
%     from maxima. This partitions the edges (into various colored
%     sets), and each vertex is incident to at most $2$ colors.
%     \item For each color, construct a (binomial) max-heap
%     indexed by height of all critical points touched by that color.
    \item Construct a painting of $\MM$ by a descending BFS from each maximum. 
    This partitions the edges into colored sets, 
    and assigns each vertex to the color classes of the edges that touch it.
    \item For each color, construct a (binomial) max-heap of the critical points touched by that color, 
          indexed by the heights of the critical points.
    \item Initialize the join tree to consists of a set of isolated vertices, one for each color class.\\
          (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
    \item Repeat until all heaps are empty:
    \begin{compactenum}
        \item Find a critical point $w$ that is at the top of all heaps containing $w$.
        \item Add $w$ to the join tree by connecting it to the lowest vertex 
              of each component of the join tree that contains a maximum whose color matches one of the heaps containing $w$.
        \item Delete $w$ from all these heaps, and merge them.
    \end{compactenum}
\end{compactenum}

@@

\subsection{Running Time}
\label{sec:runTime}

@@

\newcommand{\pruned}{\cU'}
\subsection{Paying for Pruned Shrubs: the root is everything} \label{sec:weight}

@@

\subsection{Our Main Result}
\label{sec:implications}

@@

\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
Note that the form of the theorem statements in this section differ from \Thm{main-lb}, 
as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
is an immediate corollary of the final theorem of this section, \Thm{path-main-lb}.

\subsection{Join Trees}

@@

\subsection{Contour Trees}
We first generalize previous terms to the case of contour trees.
In this section $T$ will denote an arbitrary contour tree with every internal vertex of degree $3$.
%and the height function defined over the vertex set is such that every internal vertex in either a split or join 
%(i.e.\ at most $2$ neighbors lie above or below it).

@@

\myparagraph{Acknowledgements.}
We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy's National
Nuclear Security Administration under contract DE-AC04-94AL85000.

>>>>

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time, where $n$ is the number of vertices of the mesh.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(N+\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree and $N$ is the total complexity of $\MM$.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ is a linear interpolant over distinct valued vertices, with Morse degree at most $3$. 
The degree assumption can be achieved via vertex unfolding, as discussed later,
%We define this formally later, but it suffices to think of this as a general position assumption.
and is standard in this literature~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\subsection{Previous work}

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalyi \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

@@

We detail the basic definitions about contour trees, following the terminology of Chapter 6 of Carr's thesis \cite{c-tmi-04}.
All our assumptions and definitions
are standard for results in this area, though there is some variability in notation.
The input is a continuous piecewise-linear function $f:\MM \to \RR$, where $\MM$ is a simply connected and fully triangulated simplicial complex of a topological ball in $\RR^d$,
except for specially designated \emph{boundary facets}. We assume that all pairs of vertices from $\MM$ have distinct function values, except for pairs from the same boundary facet,
and that $f$ is only explicitly defined on the vertices, and all other values are obtained by linear interpolation.

@@

\subsection{Some remarks}
{\bf 1)}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.\\

\noindent{\bf 2)}
As Carr points out in Chapter 6 of his thesis, the term contour tree can be used for a family of related structures.
Every regular vertex in $\MM$ is associated with an edge in $\reeb(\MM)$, and sometimes the
vertex is explicitly placed in $\reeb(\MM)$ (by subdividing the respective edge). This is referred
to as augmenting the contour tree, and it is common to augment $\reeb(\MM)$ with all vertices.
Alternatively, one can smooth out all vertices of up-degree and down-degree $1$ to get the
unaugmented contour tree. (For $d=2$, there are no such vertices in $\reeb(\MM)$.) 
The contour tree of \Def{tree} is the typical definition in all results on output-sensitive
contour trees.
\Thm{main-alg} is applicable for any augmentation of $\reeb(\MM)$ with a predefined
set of vertices, though we will not delve into these aspects in this paper.

@@

The intuition is that the paint spilling from the maxima in the simplicial complex, corresponds to paint spilling from the leaves in 
the join tree, which decomposes the join tree into a set of colored paths.  Unfortunately, the situation is more complex since 
while a given color class is confined to a single leaf to root path, it may \emph{not} appear contiguously on this path, as the right part of \Fig{colors} shows. 
Specifically, in this figure the far left saddle (labeled $i$) is hit by blue paint.  However, there is another 
saddle on the far right (labeled $j$) which is not hit by blue paint.  Since this far right saddle is slightly 
higher than the far left one, the merge event involving the component containing the blue mound (and also the yellow and red mounds) and corresponding to the far right saddle
will occur before the one corresponding to the far left saddle.  Hence, the vertices initially touched by blue are not contiguous in the join tree.
%
This non-contiguous complication along with the fact that heap sizes keep changing as color classes merge, causes the analysis to be technically challenging. 
We employ a variant of \emph{heavy path decompositions}, first used by Sleator and Tarjan for analyzing link/cut trees~\cite{st-dsdt-83}.
The final analysis charges expensive heap operations to long paths in the decomposition, resulting in the bound stated in \Thm{main-alg}.

@@

We now present the main algorithmic contribution, which is a new algorithm for computing join trees of any triangulated simplicial
complex $\MM$.  In subsequent sections we then show how to reduce computing the contour tree to computing the join tree by using contour surgery.

@@

\begin{compactenum}
%     \item Construct a painting of $\MM$ by a descending BFS 
%     from maxima. This partitions the edges (into various colored
%     sets), and each vertex is incident to at most $2$ colors.
%     \item For each color, construct a (binomial) max-heap
%     indexed by height of all critical points touched by that color.
    \item Construct a painting of $\MM$ by a descending BFS from each maximum. 
    This partitions the edges into colored sets, 
    and assigns each vertex to the color classes of the edges that touch it.
    \item For each color, construct a (binomial) max-heap of the critical points touched by that color, 
          indexed by the heights of the critical points.
    \item Initialize the join tree to consist of a set of isolated vertices, one for each color class.\\
          (We now iteratively fill in the rest of the join tree from the leaves towards the root.)
    \item Repeat until all heaps are empty:
    \begin{compactenum}
        \item Find a critical point $w$ that is at the top of all heaps containing $w$.
        \item Add $w$ to the join tree by connecting it to the lowest vertex 
              of each component of the join tree that contains a maximum whose color matches one of the heaps containing $w$.
        \item Delete $w$ from all these heaps, and merge them.
    \end{compactenum}
\end{compactenum}

@@

\subsection{Running time}
\label{sec:runTime}

@@

\newcommand{\pruned}{\cU'}
\subsection{Paying for pruned shrubs: the root is everything} \label{sec:weight}

@@

\subsection{Our main result}
\label{sec:implications}

@@

\section{Lower bound by path decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
Note that the form of the theorem statements in this section differ from \Thm{main-lb}, 
as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
is an immediate corollary of the final theorem of this section, \Thm{path-main-lb}.

\subsection{Join trees}

@@

\subsection{Contour trees}
We first generalize previous terms to the case of contour trees.
In this section $T$ will denote an arbitrary contour tree with every internal vertex of degree $3$.
%and the height function defined over the vertex set is such that every internal vertex in either a split or join 
%(i.e.\ at most $2$ neighbors lie above or below it).

@@

\myparagraph{Acknowledgements.}
We thank Hsien-Chih Chang, Jeff Erickson, and Yusu Wang for numerous useful discussions. This work is
supported by the Laboratory Directed Research and Development
(LDRD) program of Sandia National Laboratories. Sandia National
Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed
Martin Corporation, for the U.S. Department of Energy's National
Nuclear Security Administration under contract DE-AC04-94AL85000.
Work on this paper by Benjamin Raichel was also partially supported
by NSF CRII Award 1566137.
